---
output: html_document
editor_options: 
  chunk_output_type: console
---
## Term Project Assignment 2 - 10 pnts

In this assignment you will develop the main goal and objectives of your term project and explore, evaluate, and select data sources. A key component of this process is formulating a research question or develop a workflow. The workflow for this assignment will help you refine your question and identify the datasets needed to address it. 

### Brainstorming

**Q1. What is the primary research question/objective for your term project? (1-2 sentences)(1 pt)** <br>

ANSWER: How has variability in stream discharge, precipitation, and air temperature impacted salmon returns in Western Washington.

**Q2. What are 2-3 types of data your research question requires? Address each sub-question (2 pts)**
ANSWER:
1) For this project the variables and indicators we are looking for is salmon return data (specifically from streams with USGS gages), stream discharge data, precipitation data, and air temperature data.

2) The data should all be in the form of time-series data, however, we will likely use the map function previously used in labs to display where the stream gages we are using are located. 

3) The salmon return data will be survey type data from the stateofsalmon.wa.gov website. The stream discharge data will come from three different USGS gages. The precipitation and air temperature data will come from daymet data.

4) As salmon have a 3-5 year life cycle depending on the species a longer temporal length would be best, ideally spanning 20+ years so that patterns across multiple generations can be examined. The resolution of the data will depend on the variable. Salmon return data are reported as annual totals, so this data will be represented as a single value each year. However, using daily data for stream flow, precipitation, and air temperature would allow for examination of annual patterns and help to determine if there is a correlation between changes in these factors and salmon returns. Therefore the temporal length for these three variables will be on a daily basis. 

### Data exploration and selection

**Q3. Explore your data, this part requires 2 answers (3 pnts total)**
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install and load needed packages

```{r}
library(dataRetrieval)
library(tidyverse)
library(dplyr)
library(patchwork)
library(zoo)
library(daymetr)
library(leaflet)
library(fs)
library(ggpubr)
```

**Write a brief summary (3-4 sentences) describing your data.** 

Answer: For our project we are using four different types of data, salmon return data, stream discharge, precipitation, and air temperature. The salmon return data is from the stateofsalmon.wa.org website and is reported as one value each year of the total number of wild and hatchery fish that returned to a given river each year. The stream discharge data comes from three USGS gauges and we are looking at daily data from 1987-2023. The precipitation and air temperature data were retrieved from daymet and is unique for all three of the sites.

Generate plots:
Create at least one plot that summarizes the data and describe it's use to you. Are there gaps in the data? Does the data cover the time or space you are interested in? Are there significant outliers that need consideration? 

ANSWER BELOW PLOT

```{r}
# USGS GAGE INFO/DATA 

# This efficiently automates the process of getting all the stream data desired into one df.

gage_ids <- 
  tribble(
    ~site_name,        ~site_no,
    "Dungeness_Gage",  "12048000",
    "Elwha_Gage",      "12045500",
    "Skokomish_Gage",  "12061500")

# set all the parameters
start <- date("1987-10-01")
end <- date("2023-9-30")
para <-c("00060", "00065")

# if sites in gage_ids are changed, uncomment the code below and run it once. This stores the data locally and eventually in the repo if its pushed. This is mostly to be polite to the USGS hosting service.

# NWIS_data <- readNWISdv(gage_ids$site_no, para, start, end)
# saveRDS(NWIS_data, "NWIS_data.rdata")

# NWIS_meta <- readNWISsite(gage_ids$site_no)
# saveRDS(NWIS_meta, "NWIS_meta.rdata")

gage_ids <- 
  readRDS("NWIS_meta.rdata") %>% 
  select(site_no,
         dec_lat_va, 
         dec_long_va) %>%
  inner_join(gage_ids, .)

gage_data <-
  readRDS("NWIS_data.rdata") %>%
  inner_join(gage_ids, .) %>%
  renameNWISColumns() %>%
  group_by(site_name, site_no)


# Shows all unique grouped elements in gage_data. Verifies that all the desired gages were selected and are in the data frame.

summarize(gage_data)

```


```{r}
# WEATHER DATA ACQUISITION: This code retrieves the desired precipitation and temperature data.

# write_csv(x = select(gage_ids, -"site_no"), file = "gage_ids.csv")

# daymet_all <- download_daymet_batch(file_location = "./gage_ids.csv", start = year(start), end = year(end), internal = T, simplify = F)
# saveRDS(daymet_all, "DAYMET_all.rdata")


# Downloading data works. Now to append it...

daymet_reorg <-
  function(f) {
    bind_cols(data = f$data, site = f$site, .name_repair = "unique")
  } 

gage_data_all <- 
  lapply(readRDS("./DAYMET_all.rdata"), daymet_reorg) %>%
  bind_rows() %>%
  rename(doy = yday, 
         ann = year,
         prcp_daily_mm = prcp..mm.day.,
         t_max_C = tmax..deg.c.,
         t_min_C = tmin..deg.c.) %>%
  mutate(
    Date = make_datetime(year = ann, day = doy),
    t_avg_C = (t_max_C + t_min_C)/2) %>%
  select(site,
         Date,
         prcp_daily_mm,
         t_max_C,
         t_min_C,
         t_avg_C,
         ann
  ) %>%
  left_join(ungroup(gage_data), ., join_by(site_name == site, Date == Date))

#Now the discharge, precipitation, and temperature data is all in one data frame.
```


```{r}
# ADD FISH DATA: This code reads in csv files containing salmon return data downloaded for the particular rivers of interest from the stateofsalmon.wa.gov site.

fish_files <-
  dir_ls("./Salmon_Data/")

gage_fish_all <-
  fish_files %>%
  lapply(read_csv) %>%
  bind_rows()%>%
  filter(data_type == "TSAEJ") %>%
  mutate(site_name = paste0(str_extract(population_name, "^\\w+"), "_Gage"), .before = "stock_number") %>%
  select(abundance_qty,
         year,
         site_name) %>%
  left_join(gage_data_all,., by = c(
    "ann" = "year",
    "site_name" = "site_name"
  )
  )

# REVEL IN ITS GLORY. FISH SCRIPT REIGNS SUPREME

gage_fish_all %>%
  ggplot(aes(x = Date, y = Flow, color = site_name)) +
  geom_line() +
  facet_wrap(~ site_name, nrow  = 3, scale = "free")

```


Generate a histogram or density plot of at least one variable in your dataset. The script here will help start a density plot showing multiple variables. You may adapt or change this as needed. 

```{r}
gage_fish_all %>%
  ggplot(aes(x = abundance_qty)) +
  geom_histogram() +
  facet_wrap(~site_name, nrow = 3, scale = "free")
```

**What does the above plot tell you about the distribution of your data? Is it as expected? Why do we need to consider the distribution of data when deciding on analysis methods?**



**Q4. Putting it all together (3 pts)**
Create a table in R and export it as a .csv file to submit with this assignment .rmd.

In this table, include all datasets (up to now) that you will use for your term project. The script below specifies all of the sections required in this table, though you will need to change names and information accordingly in c(). Also, not everything in this table may pertain to you and your project. So we can change things accordingly. 

```{r, message=FALSE}
write.csv(slice(gage_fish_all, c(1:4)), "all_data_preview.csv", )

data_slice <- read.csv("./all_data_preview.csv")
```

Once this is exported, you can format if desired for readibility and submit the .csv with a completed .rmd. 

**Q5. What is your updated term project question/objective? How, specifically, will the data sources listed above help you to answer that question? (1 pt)**

How have fluctuations in stream discharge, precipitation, and air temperature influenced the survivability and return rates of Chinook salmon in Western Washington from 1987 to 2023? The data sources above will help to answer our question by providing additional insights into potential variables impacted salmon runs. Precipitation and stream discharge values will help us to analyze the flow conditions that salmon populations were experiencing within a given year and track the correlating impacts on salmon return numbers. Additionally, air temperature will help us to better understand stream conditions that the salmon were experiencing within a given season and tie these values to low flows. These climatological trends can then we tracked to analyze how variables that salmon experienced during the juvenile stage impact the return numbers three to five years later. 

